---
title: "SWIRLSS package"
author: "<a href= http://henokuga@gmail.com>Henok G. Woldu (PhD) and Kevin K. Dobbin (PhD) </a>"
date: "`r Sys.Date()`"
output: 
   
   rmarkdown::html_vignette:
   self_contained: no
   fig_caption: yes
   css: SWIRLSSvignettestyle.css
vignette: >
  %\VignetteIndexEntry{SWIRLSS}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

# Summary
Advancement in genom sequencing in the last few years have lead to a surge of predictive biomarker discoveries. Parallel to that, statistical methods for evaluating these biomarkers have been advancing. A metric $\Theta$ developed by [Janes et.al (2014)](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4626364/), is now widely used as  a measure of the biomarker's clinical 
benefit for patients. This metric measures the decrease in the expected event rate as a result of a biomarker guided theraphy. However, there is no smaple size calculation method developed that can help to design  a study which is enough powered  in order to evaluate the biomarker's clinical utlity. The _Squared Width Inversion Linear Regression Sample Size (SWIRLSS)_ is a package used to estimate a sample size $n$ such that the 95\% CI mean width of the metric $\Theta$ is smaller than a user defined length ($W_{targ}$).

## Package description

The **SWIRLSS**  package consisits of several functions used to estimate the sample size $n$ so that the mean width of the 95\% confidence interval for $\Theta$ is less than a user defined length $(W_{targ})$. The main functions of the package are:

1. **coeffgen** : A function used to change the clinician defined values to model coefficients.

2. **logdata** : A function  used to generate data from a logistic regression model given by 
$$Ln\left[ \frac{Pr(Y = 1|T,X)}{1-Pr(Y = 1 | T,X)} \right] = \beta_0 +\ \beta_1X+\beta_2T+\beta_3 T*X$$
 Where $Y$ is the outcome which is binary, $T$ is the treatment assignment ( $T = 1$ if treated and $T = 0$ if not treated) and $X$ is a continous biomarker measured at base line for a each subject.
 
3. **eventprob** : A function used to estimate the probability of an event under the three different treatment options. 

   (a) Treat All: estimates the probility of an event when all subjects are assigned to treatment group.
   (b) Treat None: estiamtes the probability of an event when all subjects are assigned to no treament (control) group.
  (c) Optimal Treament: estiamtes the probability of an event when treatment assignment is made based on the biomarker
    value of each subjects. That is, a subject will be assigned to treament if he/she will benefit from it and to no treament
    otherwise.
    
4. **bootsmethod** : A function used to get the converage probability of $\Theta$.


5. **ssswirl** : A function used to estimate the sample size $n$ that ensures the 95\% CI mean width of $\Theta$ is less than 
   a defined length $(W_{targ})$



## Using this package

The main steps to follow to use and/or extend this package are as follows:


### Install the package

It is not yet on CRAN, but once it is available it can be installed and used as

```{r, eval=FALSE , echo=T}
 
install.packages("SWIRLSS")  # install it first 
   
library(SWIRLSS)             # load the package 

```


The very updated version of this package can be obtained from my _github_ account [Henok G. Woldu] (https://github.com/henok535). To install it from my github, the `devtools` package has to be  first installed. Here is how you can install it directly from github:

```{r, eval=FALSE, echo=TRUE}

install.packages("devtools")  # install the devtools package if you have not done it before.

library(devtools)            # load the package devtools

# to install the SWIRLSS from my github acount ("henok535/SWIRLSS")

install_github("henok535/SWIRLSS")

library(SWIRLSS)     
 
# Well come to SWIRLSS package. It is now ready for use!

```


## Examples for using each function 

########## Required packages ######

```{r}

library(boot)

```


### Changing Clinician Inputs to Model Coefficients


```{r}

# Here are some possible user defined values ( 11, k2, k3 and k3). These can be adjusted to meet specific
# problem under consideration.

kval1 <- c(K1 = log(0.25/0.75), k2 = log(0.75/0.25), k3 = log(0.75/0.25), k4= log(0.25/0.75))
kval2 <- c(k1 = log(0.10/0.90), k2 = log(0.90/0.10), k3 = log(0.90/0.10), k4 = log(0.10/0.90))
kval3 <- c(k1 = log(0.10/0.90), k2 = log(0.55/0.45), k3 = log(0.90/0.10), k4 = log(0.45/0.55))
kval4 <- c(k1 = log(0.25/0.75), k2 = log(0.75/0.25), k3 = log(0.50/0.50), k4 = log(0.50/0.50))
kval5 <- c(k1 = log(0.90/0.10), k2 = log(0.45/0.55), k3 = log(0.10/0.90), k4 = log(0.55/0.45))
kval6 <- c(k1 = log(0.60/0.40), k2 = log(0.40/0.60), k3 = log(0.50/0.50), k4 = log(0.50/0.50))


# get the 25th and 75th percentile value of the biomarker value  assuming it is
# is normaly or uniformly distributed

set.seed(535)
pardist <- c(0,1)    # standard uniform / normal. We can change this one as needed

bmrkunif <- runif(100000,pardist[1],pardist[2]) # Assuming the biomarker has uniform distribution
bmrknorm <- rnorm(100000,pardist[1],pardist[2]) # Assuming the biomarker has normal distribution

zval <- c(quantile(bmrkunif)[2], quantile(bmrkunif)[4]) #25th and 75th percentile biomarker

modcoeff <- coeffgen(kval1, zval)
modcoeff    # this provideds the coefficients beta0, beta1, beta2 and beta3
            # which will be used to generate data from the logistic model specified above.
```


## Calculate the event probability under each scenario

```{r}

# When the biomarker distribution is uniform

Event_None <- eventprob(modpar = modcoeff, dist = "uniform", trtopt = "treat_none",  bmrkpar = pardist)
Event_All <- eventprob(modpar = modcoeff, dist = "uniform", trtopt = "treat_all",  bmrkpar = pardist)
Event_Opt <- eventprob(modpar = modcoeff, dist = "uniform",  trtopt = "treat_opt",  bmrkpar = pardist)

# then Theta1 and Theta0 will be

Theta1_true <- Event_All - Event_Opt
Theta0_true <- Event_None - Event_Opt
c(Theta1_true, Theta0_true)

# When the biomarker is generated from a normal distribution, one can uncomment the following codes
# to calculate the event probability under each scenario the their respective thetas.

  #Event_None <- eventprob(modpar = modcoeff, dist = "normal", trtopt = "treat_none",  bmrkpar = pardist)
  #Event_All <- eventprob(modpar = modcoeff, dist = "normal", trtopt = "treat_all",  bmrkpar = pardist)
  #Event_Opt <- eventprob(modpar = modcoeff, dist = "normal", trtopt = "treat_opt",  bmrkpar = pardist)

# then Theta1 and Theta0 will be

    #Theta1_true <- Event_All - Event_Opt
    #Theta0_true <- Event_None - Event_Opt
    #c(Theta1_true, Theta0_true)

```



## Generate a data from a logistic model

```{r}

m = 25      # number of data sets
n = 500     # sample size for each data set


mydat <- logdata(m, n, modcoeff, "uniform", pardist)

datt <- mydat[[1]]     # sample of one data set            

```


## Use bootstrap to get the coverage probability and 95\% CI mean width for Theta.

When running this, it takes few minutes. If the objective is just to understand the 
program, one can run this with few data sets ( 2 to 5) and small sample size ( 100 to 300). 
That is, change m and n in the previous step.


#### Please uncomment all the following codes to run the program.

```{r}

# alldatout <- lapply(mydat, function(a) bootsmethod(a, "uniform", pardist))
# outdat <- as.data.frame(matrix(unlist(alldatout), ncol = 11, byrow = T))
# 
# Finaloutput <- list(Treat_None = mean(outdat$V1),
#                     Treat_All = mean(outdat$V2),
#                     Treat_Opt = mean(outdat$V3),
#                     Theta1 = mean(outdat$V4),
#                     Theta195CI = c(mean(outdat$V6), mean(outdat$V7)),
#                     width_Theta1 = mean(outdat$V5),
#                     Converage_Theta1 = length(intersect(which(outdat$V6 < Theta1_true),
#                                                 which(outdat$V7 > Theta1_true))),
#                     Theta0 = mean(outdat$V8),
#                     Theta095CI = c(mean(outdat$V10), mean(outdat$V11)),
#                     width_Theta0 = mean(outdat$V9),
#                     Converage_Theta0 = length(intersect(which(outdat$V10 < Theta0_true),
#                                             which(outdat$V11 > Theta0_true)))
#                   )
# 
# Finaloutput


```


## Sample Size Estimation

Similarly, running this funciton takes few minutes . User who just want to familarize with the program
can change the mcdat to fewer squence of number each representing a sample size n and run the program.


#### uncomment this code to run it

```{r}

# newdat1 <- data.frame(w = c(0.10,0.15,0.20)^-2)
# mcdat = seq(100,1500,50)
# samplesize1 <- ssswirl(modcoeff, "uniform", mcdat, m = 10, pardist,
#                                    bmrkdgn = "strategy", newdat1)
# 
# 
# samplesize1

```








